Expression QTL hotspot dissection with mediation analysis and pleiotropy testing

### 2019-07-29 analysis/paper/hotspot-one-chromosome.Rmd
fixed equation labels

```{r}
tibble(symbol = rownames(pm[indic_nonlocal, indic_nonlocal])[ord_nonlocal], index = 1:length(symbol)) %>%
  left_join(nonlocal_annot, by = "symbol") %>%
  ggplot() + geom_point(aes(y = pos, x = index))
```

>>>>

```{r}
bar <- foo %>%
  filter(local_symbol == keller_mediator)

p <- tibble(nonlocal_symbol = rownames(pm[indic_nonlocal, indic_nonlocal])[ord_nonlocal], index = 1:length(nonlocal_symbol)) %>%
  inner_join(bar, by = "nonlocal_symbol") %>% 
  left_join(nonlocal_annot, by = c("nonlocal_symbol" = "symbol")) %>%
  ggplot() + geom_point(aes(y = pos, x = index, color = lod_diff > 1.5, nonlocal = nonlocal_symbol, lod_diff = lod_diff))

ggplotly(p)
```

### 2019-07-29 analysis/paper/hotspots-paper.Rmd
fixed equation labels

---
title: "Expression QTL hotspot dissection with mediation analysis and pleiotropy testing"
author: "Frederick J. Boehm, Brian S. Yandell, and Karl W. Broman"
date: "7/27/2019"
bibliography: research.bib
numberSections: true
autoEqnLabels: true
output: 
  bookdown::html_document2:
    pandoc_args: 
      - --filter
      - /usr/local/Cellar/pandoc-crossref/0.3.4.1/bin/pandoc-crossref
---

@@

@@

>>>>

---
title: "Expression QTL hotspot dissection with mediation analysis and pleiotropy testing"
author: "Frederick J. Boehm, Brian S. Yandell, and Karl W. Broman"
date: "7/27/2019"
bibliography: research.bib
numberSections: true
linkReferences: true
link-citations: true
output: 
  bookdown::html_document2:
    pandoc_args: 
      - --filter
      - /usr/local/Cellar/pandoc-crossref/0.3.4.1/bin/pandoc-crossref
---

@@

```{r, tmem189-scatter, fig.cap="Colors correspond to mediation by Tmem189"}
bar <- foo %>%
  filter(local_symbol == "Tmem189")

p <- tibble(nonlocal_symbol = rownames(pm[indic_nonlocal, indic_nonlocal])[ord_nonlocal], index = 1:length(nonlocal_symbol)) %>%
  inner_join(bar, by = "nonlocal_symbol") %>% 
  left_join(nonlocal_annot, by = c("nonlocal_symbol" = "symbol")) %>%
  ggplot() + geom_point(aes(y = pos, x = index, color = lod_diff > 1.5, nonlocal = nonlocal_symbol, lod_diff = lod_diff))

ggplotly(p)
```

## Chr 2 hotspot results

@@

```{r child="hotspot-one-chromosome.Rmd", eval = FALSE}
```

### 2019-07-29 analysis/paper/intro.Rmd
fixed equation labels

Genetics studies in model organisms like mice can identify
genomic regions that affect complex traits, such as systolic blood pressure and body weight [@sax1923association;@soller1976power;@lander1989mapping;@broman2009guide;@jansen2007quantitative]. 
These genomic regions are called "quantitative trait loci" or "QTL".
A genome-wide QTL "scan" reveals
associations between genotypes and phenotypes by considering each position along the genome,
one at a time, as a candidate QTL for the trait of interest.
A region with strong evidence of association with a complex trait, then, defines a QTL (for that trait).
Because nearby markers have correlated
genotypes, a QTL in a two-parent cross often spans multiple megabases in length
and may contain more than a hundred genes.
Identification of a causal gene (for a given complex trait) from among those genes near the QTL is
challenging and may require costly and time-consuming experiments.
The growing need for greater QTL mapping resolution fueled development over the
last two decades of model organism multiparental populations for high-resolution QTL mapping [@de2017back;@churchill2004collaborative;@svenson2012high;@huang2012multiparent;@shivakumar2018soybean;@huang2011analysis;@kover2009multiparent;@tisne2017identification;@stanley2017genetic].
With experimentalists now measuring tens of thousands of biomolecular traits
in multiparental populations, the systems genetics community needs multivariate
statistical tools to fully examine the large volumes of data [@keller2018genetic;@chick2016defining].
Identifying loci that affect multiple traits can aid in identifying biomolecular interactions and in clarifying complex trait genetic architecture. A QTL that affects more than one trait is called a "pleiotropic" QTL.
A test of pleiotropy vs. separate QTL is one multivariate statistical tool that will inform
complex trait genetics by enabling researchers to identify the number of
unique QTL in a genomic region of interest.

>>>>

Genetics studies in model organisms like mice can identify
genomic regions that affect complex quantitative traits, such as systolic blood pressure, body weight, and, more recently, biomolecular traits such as gene expression levels and protein concentrations [@sax1923association;@soller1976power;@lander1989mapping;@broman2009guide;@jansen2007quantitative;@chick2016defining]. 
These genomic regions are called "quantitative trait loci" or "QTL".
A genome-wide QTL "scan" reveals
associations between genotypes and phenotypes by considering each position along the genome,
one at a time, as a candidate QTL for the trait of interest.
A region with strong evidence of association with a complex quantitative trait, then, defines a QTL (for that trait).
Because nearby markers have correlated
genotypes, a QTL in a two-parent cross often spans multiple megabases in length
and may contain more than a hundred genes.
Identification of a causal gene (for a given complex trait) from among those genes near the QTL is
challenging and may require costly and time-consuming experiments.
The growing need for greater QTL mapping resolution fueled development over the
last two decades of model organism multiparental populations for high-resolution QTL mapping [@de2017back;@churchill2004collaborative;@svenson2012high;@huang2012multiparent;@shivakumar2018soybean;@huang2011analysis;@kover2009multiparent;@tisne2017identification;@stanley2017genetic].

Recent systems genetics studies have sought to identify QTL for thousands of biomolecular traits [@keller2018genetic;@chick2016defining].

### 2019-07-29 analysis/paper/methods.Rmd
fixed equation labels

For each hotspot, we tested all pairs of traits, regardless of whether the traits were local or nonlocal. We performed a
two-dimensional, two-QTL scan over a genomic region that included the entirety of the
hotspot. We fitted bivariate linear mixed effects models at each ordered pair of markers in a two-dimensional grid [@boehm2019testing].

$$
vec(Y) = X vec(B) + vec(G) + vec(E)
$${eq:mvlmm}

$$
G \sim MN_{n x 2}(0, K, V_g)
$$

$$
E \sim MN_{n x 2}(0, I_n, V_e)
$$

@@

$$
LOD = \log_{10} \left(\frac{\max_{\lambda_1, \lambda_2}L(B, \Sigma, \lambda_1, \lambda_2)}{\max_{\lambda}L(B, \Sigma, \lambda, \lambda)} \right)
$${eq:lod}

@@

$$
Y = \alpha 1 + WC + E
$${eq:model1}

$$
Y = XB + WC + E
$${eq:model2}

$$
Y = \alpha 1 + WC + M\beta + E
$${eq:model3}

$$
Y = XB + WC + M\beta + E
$${eq:model4}

In all of our mediation analyses, the nonlocal transcripts are the targets, while the local
transcripts serve as candidate mediators of the QTL-target relationship.
In Equations~\ref{model1} through \ref{model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator, $\beta$ is the effect
of the mediator on the complex trait, 1 is a vector with all entries set to 1, $\alpha$
is the mean trait value, and $E$ is a vector of normally distributed random errors.

>>>>

$$
vec(Y) = X vec(B) + vec(G) + vec(E)
$${#eq:mvlmm}

For each hotspot, we tested all pairs of traits, regardless of whether the traits were local or nonlocal. We performed a
two-dimensional, two-QTL scan over a genomic region that included the entirety of the
hotspot. We fitted bivariate linear mixed effects models (Equations [-@eq:mvlmm;-@eq:g;-@eq:e]) at each ordered pair of markers in a two-dimensional grid [@boehm2019testing].

$$
G \sim MN_{n x 2}(0, K, V_g)
$${#eq:g}

$$
E \sim MN_{n x 2}(0, I_n, V_e)
$${#eq:e}

@@

$$
LOD = \log_{10} \left(\frac{\max_{\lambda_1, \lambda_2}L(B, \Sigma, \lambda_1, \lambda_2)}{\max_{\lambda}L(B, \Sigma, \lambda, \lambda)} \right)
$${#eq:lod}

@@

$$
Y = \alpha 1 + WC + E
$${#eq:model1}

$$
Y = XB + WC + E
$${#eq:model2}

$$
Y = \alpha 1 + WC + M\beta + E
$${#eq:model3}

$$
Y = XB + WC + M\beta + E
$${#eq:model4}

In all of our mediation analyses, the nonlocal transcripts are the targets, while the local
transcripts serve as candidate mediators of the QTL-target relationship. We only considered as potential mediators those local traits with strong univariate LODs.
In Equations~\ref{model1} through \ref{model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator, $\beta$ is the effect
of the mediator on the complex trait, 1 is a vector with all entries set to 1, $\alpha$
is the mean trait value, and $E$ is a vector of normally distributed random errors.
